---
title: "final"
author: "Manik Mukherjee"
due date: "4/25/3030"
output: html_notebook
---

```{r}
#import libraries to make the rest of the code work
library(DataComputing)
```

### Research Question

# Do county election results correlate to the percentage of the population that is not white?

### Data Access

The link to the third source is required to be linked to according to the download website: https://simplemaps.com/data/us-zips

# Import the data
```{r}
primary <- read.csv(file = "countypres_2000-2016.csv")
secondary <- ZipDemography[, c("Totalpopulation", "White", "ZIP")]
third <- read.csv(file = "uszips.csv")[, c("zip", "county_name")]
```

# Inspect the data
```{r}
head(primary)
str(primary)
head(secondary)
str(secondary)
head(third)
str(third)
```


### Data Wrangling

# General Data Wrangling

```{r}
# find the total number of votes for democrats ever
democrat <- primary %>%
  filter(party == "democrat") %>%
  summarize(total_democrats = sum(totalvotes))
democrat

# find the total number of votes for republicans
republican <- primary %>%
  filter(party == "republican") %>%
  summarize(total_republicans = sum(totalvotes))
republican
```

# Join Operation

```{r}
third[, "county_name"] <- as.character(third[, "county_name"])
primary[, "county"] <- as.character(primary[, "county"])
secondary[, "ZIP"] <- as.numeric(secondary[, "ZIP"])

names(secondary)[3] <- "zip"
names(third)[2] <- "county"
```


```{r}
combined <- left_join(secondary, third)
combined <- left_join(primary, combined)
head(combined)
```

# Regular Expressions

This regular expression extracts the last name (if applicable) from the candidate name, as many people refer to presidents by just their last name.

```{r}
last_names <- primary %>%
  select(candidate) %>%
  extractMatches("([^ ]+$)", candidate)
head(last_names)
```

This filters out counties that have "son", "ger", or "gua" in their name

```{r}
patterns <- third %>%
  filter(grepl("son|ger|gua", county, ignore.case = TRUE))
head(patterns)
```

# User Defined Functions

this function will print popular if the input over 10,000 votes and not popular if the input is not.  This could be used in the context of votes

```{r}
popular <- function(votes) {
  if (votes > 10000) {
    status <- "popular"
  } else {
      status <- "not popular"
    }
  status
}
```

this function will generate a random sample of size n from any dataset so so long as the sample size isn't greater than the size of the dataset.

```{r}
random_sample <- function(dataset, n) {
  total <- 1:nrow(dataset)
  random <- sample(total, size = n, replace = FALSE)
  dataset[random, ]
}

random_sample(primary, 10)
``` 

# Loops or Control Flow

This while loop finds the amount of rows in primary that are in Alabama until it moves on to another state (there are more rows in the state of Alabama later). 

```{r}
row <- 1
while (primary[row, "state"] == "Alabama") {
  row <- row + 1
}
row
  
``` 

This for loop finds the amount of people that have voted for a non democrat or republican candidate

```{r}
n <- 0
for (x in 1:nrow(primary)) {
  if (primary[x, "candidate"] == "Other") {
    n <- n + 1
  }
}
n
``` 


### Data Visualization

This is data wrangling for the next plot

```{r}

winner <- primary %>%
  filter(year == 2000) %>%
  distinct(state, county, party, candidatevotes) %>%
  group_by(state, county) %>%
  filter(candidatevotes == max(candidatevotes))

white_prop <- combined %>%
  filter(year == 2000) %>%
  group_by(state, county) %>%
  summarise(Totalpopulation = sum(Totalpopulation, na.rm = TRUE), White = sum(White, na.rm = TRUE))

winner_white <- left_join(winner, white_prop, by = c("state","county")) %>%
  mutate(prop = White/Totalpopulation) %>%
  group_by(party, prop=cut(prop, breaks= seq(0, 1, by = 0.1))) %>% 
  summarise(n= n())

``` 

This shows a bar plot of the the the number of counties that chose a specific party versus the percentage of white people living in that county.

```{r}

ggplot(data=winner_white,aes(x=prop,y=n ,fill=party)) + geom_bar(stat='identity',position='dodge', width=.9) +
  scale_fill_manual("legend", values = c("democrat" = "blue", "republican" = "red"))

``` 

This shows the percentage of counties that voted democrat and republican given the percentage of white people in the county.

```{r}
percentages <- left_join(winner, white_prop, by = c("state", "county")) %>%
  mutate(prop = White/Totalpopulation) %>%
  group_by(party, prop=cut(prop, breaks= seq(0, 1, by = 0.1))) %>% 
  summarise(n= n()) %>%
  group_by(prop) %>%
  mutate(percent = n/sum(n))
  

ggplot(data=percentages,aes(x=prop,y=percent)) + 
  geom_point(aes(color=party)) +
  scale_color_manual(values=c('#0000FF','#FF0000'))

``` 

This shows the overall distribution of the percentage of white people in counties in the US.

```{r}
white_prop <- white_prop %>%
  mutate(prop = White/Totalpopulation)
  

ggplot(white_prop, aes(x=prop)) +
  geom_boxplot()

``` 
